---
layout: post
title: Content Management - First post with RMarkdown
series: Content Strategy
published: true
math_symbols: false
categories: [blogging, web design, web development]
tags: [web design, content management, resources, reproducible research]
---

## Introduction

At the end of the [previous post]({{ site.baseurl }}{% post_url 2017-03-17-content-management-blogging_with_R %}) we went through a diverse range of some of the ways to incorporate the RMarkdown content into my website. The path that I settled on was [knitr-jekyll](https://github.com/yihui/knitr), since it looked fairly simple to import into my workflow.

## Result: Cooking with gas!

Now that we are pretty much setup, let's see how it all works on some simple data. Below is the first chunk in our analysis, which sets things up for us. We will be looking at the **`pressure`** dataset, which is included in the datasets package that comes with the core distribution of R.

In this case, the setup is pretty simple and self explanatory.

```{r, setup}
# obtain the dataset
data(pressure)
# display the first couple of rows
head(pressure)
```

## Displaying tabular data

I have used the **kable()** function from knitr to good effect in the past to create some nice tables. Typically, I don't bother specifying any other arguments besides the dataset, but today we will dabble a little :smile:.

```{r, kableTable}
# displaying the same dataset using knitr's kable() function to format it nicely.
# base caption with table description.
pressureDataCaption = "Brief Snapshot of the Pressure dataset"

# constructing table with modified caption
# optionally, we could specify a format such as pandoc. see ?knitr.kable
knitr::kable(head(pressure), format = "html", caption = paste(pressureDataCaption, "with kable() from the knitr package."))
```

Alternatively, one could use the **[pander](https://cran.r-project.org/web/packages/pander/README.html)** package. I found an example of this while looking at the [SciViews](http://www.sciviews.org/blog/The-SciViews-Jekyll-Template/) website, which interestinly uses Jekyll and servr :smile:.

```{r, panderTable}
# alternatively, one could use the pander() function from pander.
require(pander)
pander::pander(head(pressure), caption = paste(pressureDataCaption, "with pander() from the pander package."), style = "rmarkdown")
```

RStudio formats the table content a little better, but this will do nicely... need to work on that a bit later on though.

## Displaying plots

Now for the really interesting part... plotting graphics :smile:. This section contains a simple comparison of the base plotting system with that of ggplot2. This is probably the most challenging thing about knitting documents in R, because: 

* the figures need to be generated and stored in the correct location
* the figures need to be referred to properly within the created document

Needless to say, this was an exciting development when this worked :smile:

Note: I randomly found this stackoverflow question on [embedding existing graphics in knitr chunks](Interesting note: http://stackoverflow.com/questions/36350213/how-to-insert-plain-picture-jpeg-png-from-chunk-with-knitr). I don't need this now but it might come in handy. Remember...

> "It is better to have it and not need it... than to need it and not have it" :wink:

### Base ploting system

An example of a pretty basic but informative plot produced using the base plotting system (R's simple default).

```{r, basic_data_plot}
# a simple plot of the data
plotTitle = "Pressure Dataset: Plot of Pressure and temperature"
plotAxes = c(x = "Temperature", y = "Pressure")
plot(pressure$temperature, pressure$pressure, main = plotTitle, 
    xlab = plotAxes[['x']], ylab = plotAxes[['y']])
```

### ggplot2 ploting system

The same information displayed using the ever popular and highly potent [ggplot2](http://ggplot2.org/).

```{r, fig.align='center', fig.height=4}
library(ggplot2)
pressureFig = ggplot(data=pressure) + aes(x=temperature, y=pressure) + geom_point(color = "blue") + theme_minimal() + labs(title = plotTitle, x = plotAxes[['x']], y = plotAxes[['y']])
pressureFig
```

## Setup

That was a nice and simple walkthrough to show how RMarkdown data can be processed and incorporated neatly into a Jekyll-based web authoring framework. More importantly, this is essentially a first test of getting knitr-jekyll integrated into my system. 

The obvious question is: "how did you do this?". I deliberatly left this to last because I wanted to demonstrate the results of the process as a justification (of sorts) for the effort invested before describing how this was achieved.

I basically followed the strategy that I discussed [previously]({{ site.baseurl }}{% post_url 2017-03-17-content-management-blogging_with_R %}#knitr-jekyll-solution), making modifications as I went. The steps to create the current setup were:

1. Install the servr R package
2. Create the **`_source/`** subdirectory for RMarkdown version of posts to process
    * included a **`_source/helper_scripts/`** subdirectory to house the helper scripts that I need
    * the first helper file `process_posts.R` contains the **`process_rmd_posts()`** function, which is basically a wrapper of **`servr::jekyll()`** with a custom jekyll command (I prefix with "bundle exec" due to my present config and I sometimes work with drafts or future posts). The most important change is that I set the `serve` parameter to false (i.e. **`serve=F`**). This was because RStudio crashes everytime I try to build using the default option of true :alien:. However, this isn't a problem for me as I don't want to serve the website in RStudio anyway and use a web browser instance instead :wink:.
3. Imported the original [build.R](https://github.com/yihui/knitr-jekyll/blob/74d652ca5641bac65f245ab866a5d901dd5116d3/build.R) code and made modifications (compare with my current version [here]()) during setup.
    * images saved to the **`images/fig/posts/`** subdirectory because I didn't want to have another "top level" subdirectory devoted to images (originally called **`figure/`**), when the `images/` folder housed all of my site graphics. Took a litle mucking around but it was worth it. This is setup so that any Rmd content with the pages layout can later be stored in **`images/fig/pages/`** (still to work on that further). 
    * Fixed an issue with figure image links where images were stored in the expected location but not properly linked in the document. I can only presume that this was due to how the `baseurl` YAML variable was setup in the Jekyll Now `_config.yaml` file.

## Conclusion

This is a really good start and I will work on dealing with processing drafts and pages later :smile:.
