---
title: 'Factorial Methods: Preprocessing for PCA'
layout: post
math_symbols: yes
series: Factorial Analysis Intro
tags:
- overview
- factorial methods
- dimension reduction
- exploratory analysis
- unsupervised learning
- PCA
categories: data science
---

## Introduction

[this tutorial](http://www.sthda.com/english/wiki/principal-component-analysis-how-to-reveal-the-most-important-variables-in-your-data-r-software-and-data-mining), with some background of the data covered [here](http://www.sthda.com/english/wiki/factominer-and-factoextra-principal-component-analysis-visualization-r-software-and-data-mining). 


they also have background on [other factorial methods](http://www.sthda.com/english/wiki/factoextra-r-package-easy-multivariate-data-analyses-and-elegant-visualization)

[FactoMineR](http://factominer.free.fr/classical-methods/principal-components-analysis.html) also have a walkthrough of this example on their website.


## Data analysis

### Analysis setup

```{r, setup_chunk, echo=FALSE}
# load packages
library("FactoMineR")
library("factoextra")
# convenience helper
prettyTable = function(...) knitr::kable(...)

# load data
data(decathlon2)

# Extract active variables/individuals for PCA
activeDecathlon2 <- decathlon2[1:23, 1:10]
```

### Data exploration

Brief

original dataset
```{r}
prettyTable(head(decathlon2))
```

active data subset
```{r}
prettyTable(head(activeDecathlon2))
```

### Data preparation

```{r}
## save as a normalisation or preprocessing metadata data.frame
dataPreproc = as.data.frame(rbind(
    mean = colMeans(activeDecathlon2),
    sd = sapply(1:length(activeDecathlon2), 
        FUN = function(x) sd(activeDecathlon2[,x]))
))

names(dataPreproc) = names(activeDecathlon2)

## Preprocessing - Standard normalisation
preprocessPCA = function(pcaData, preProcMeta){
    for(i in 1:length(pcaData)){
        pcaData[,i] = (pcaData[,i] - preProcMeta[1,i]) / preProcMeta[2,i]
    }
    return(pcaData)
}

## Apply preprocessing to active dataset
preprocDecathlon2 = preprocessPCA(pcaData = activeDecathlon2, 
    preProcMeta = dataPreproc)
```

### Run PCA

note: need to set scale.unit = False in order to apply your own preprocessing, otherwise, the data will be scaled to  unit variance as per **`?PCA`**.

```{r}
### Run PCA from factoMiner and inspect output
pcaResult <- PCA(preprocDecathlon2, scale.unit = F, graph = FALSE)
print(pcaResult)
```

## Conclusion

Next post... more details about the implementations

Then next post... preprocessing implications


